---
title: "Simple Trees"
subtitle: "Some concepts"
author: 
- Christophe Bontemps & Patrick Jonsson - SIAP
output:
  html_document:
    df_print: paged
    toc: yes
    keep_md: yes
    code_folding: show
    fig_width: 6.5
    fig_height: 4
  pdf_document:
    df_print: kable
    toc: yes
    keep_tex: yes
    fig_width: 6.5
    fig_height: 4
    extra_dependencies: ["float"]
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r Knitr_Global_Options, include=FALSE, cache=FALSE}
library(knitr)
opts_chunk$set(warning = FALSE, message = FALSE, 
               fig.pos = "!H", fig.align = "center",
               autodep = TRUE, tidy = FALSE, cache = TRUE)
#opts_chunk$set(cache.rebuild=TRUE) 

# My colors:
SIAP.color <- "#0385a8"
```

`r if(knitr:::pandoc_to() == "latex") {paste("\\large")}` 


```{r Libraries, echo=FALSE}

# Data management
library(dplyr)
library(ggplot2)
library(grid)
library(gridExtra)

# Statistical data sets
library(ISLR)
library(AppliedPredictiveModeling)

# Nice tables
library(xtable)
library(kableExtra)

# Decision tree
library(caret)
library(rpart)
library(visNetwork)

# build-in color palette
library(RColorBrewer)
library(Polychrome) 

```






```{r data}
data(twoClassData)
names(predictors) <- c("Education", "Income")
levels(classes)[levels(classes)=="Class1"] <- "Urban"
levels(classes)[levels(classes)=="Class2"] <-  "Rural"
twoClass <- data.frame(predictors,classes)
#twoClass$classes <- relevel(twoClass$classes, ref="Urban")

twoClassColor <- brewer.pal(3,'Set2')[2:1]
names(twoClassColor) <- c('Urban','Rural')
Ruralcolor <- brewer.pal(3,'Set2')[1:1]
Urbancolor <- brewer.pal(3,'Set2')[2:2]
```


```{r summary}
summary(twoClass$classes)
```

Note that we have defined *Urban* as being the first class, but we may want to change this along the way.

```{r bivariatefig}
ggplot(data = twoClass,aes(x = Education, y = Income)) + 
  geom_point(aes(color = classes), size = 1, alpha = .6) +
  scale_colour_manual(name = 'classes', 
                      values = twoClassColor) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0))+
  theme_minimal()
```


```{r vistree}
tree_fit <- rpart(classes~., data= twoClass, control = rpart::rpart.control(cp = 0.001))

Predictions <- predict(tree_fit, twoClass, type="class")
      
# confmat <- confusionMatrix(table(predicted = Predictions, Actual = twoClass$classes) )
# confmat$table

visTree(tree_fit, digits = 2, main = "Complete Decision tree")

```




```{r}
## Testing 
cp.input <- 0.01
maxdepth.input <- 5
tree_fit <- rpart(classes~., data= twoClass, 
                  control = rpart::rpart.control(cp = cp.input, 
                                                 maxdepth = maxdepth.input))

visTree(tree_fit, digits = 2, 
        main = paste("Decision tree with a depth of ", maxdepth.input, "" ))

```


Using *rpart* package one has a nice representation of the tree where
Each node shows:
- the predicted class (Urban or Rural),
- the predicted probability of survival,
- the percentage of observations in the node

```{r rparttree}
# Other option
library(rpart.plot)

rpart.plot(tree_fit, 
           main = paste("Decision tree with a max depth of ", maxdepth.input, "" ),
           type = 4,
           under = TRUE, box.palette = c(Urbancolor, Ruralcolor),
           clip.right.labs = FALSE, branch = .3)
```
On can also 

```{r rules}
rules <- rpart.rules(tree_fit)

kable(rules[,3:ncol(rules)]) 
```

One can then compute a step by step decription of the tree generated here


```{r Step0}
# Working with rules
myrules <- apply(rules, 2, function(x) gsub(",", "", gsub("([a-zA-Z]),", "\\1 ", x)))

#Original scatter plot

p <- ggplot(data = twoClass,aes(x = Education, y = Income)) + 
  geom_point(aes(color = classes), size = 1, alpha = .5) +
  scale_colour_manual(name = 'classes', 
                      values = twoClassColor) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0))+
  theme_minimal()

p + labs(title = "Position of Rural and Urban households in (Education, Income) space", 
          subtitle = paste(nrow(twoClass), "observations")
  )
```


```{r Step1}
# First node 
p1 <- p +annotate("segment", 
                  x = min(twoClass$Education),
                  xend = max(twoClass$Education),
                  y = 0.2,
                  yend = 0.2,
                  size = 1.5,
                  color = Ruralcolor ) +  
  labs(title = "First node", 
           subtitle = paste("Rule: ", gsub(",","",toString(myrules[5,2:6])))
  )
p1

p1A <- p1 + annotate("rect", 
               xmin= min(twoClass$Education),
               xmax = max(twoClass$Education),
               ymin = min(twoClass$Income),
               ymax = 0.2,
               fill = Ruralcolor, alpha = 0.2 )
p1A 

# Highlight Misclassified 
misclassU <- twoClass %>%
  filter((Income < 0.2 & classes == "Urban"))

misclassR <- twoClass %>%
  filter((Income > 0.2 & classes == "Rural"))
         
p1U <- p1A + geom_point(data = misclassU, 
                 aes(x= Education, y = Income), 
                 colour = Urbancolor, 
                 size = 2)
p1U

p1R <-  p1U + geom_point(data = misclassR, 
                 aes(x= Education, y = Income), 
                 colour = Ruralcolor, 
                 size = 2)
p1R

```


```{r Step2}
# Second node
p2 <- p1A +annotate("segment", 
                  x = 0.13,
                  xend = 0.13,
                  y = 0.2,
                  yend = max(twoClass$Income),
                  size = 1.5,
                  color = Urbancolor )+
   labs(title = "Second node:", 
           subtitle = paste("Rule: ", gsub(",","",toString(myrules[4,2:13])))
           )
p2
p2A <- p2 + annotate("rect", 
              xmin= min(twoClass$Education),
              xmax = 0.13,
              ymin = 0.2,
              ymax = max(twoClass$Income),
              fill = Ruralcolor, alpha = 0.2 )
p2A 


# Highlight Misclassified 
misclassU <- twoClass %>%
  filter((Income > 0.2 & Education < 0.13 & classes == "Urban"))

misclassR <- twoClass %>%
  filter((Income > 0.2 & Education > 0.13 & classes == "Rural"))
         
p2U <- p2A + geom_point(data = misclassU, 
                 aes(x= Education, y = Income), 
                 colour = Urbancolor, 
                 size = 2)
p2U
p2R <- p2U+  geom_point(data = misclassR, 
                 aes(x= Education, y = Income), 
                 colour = Ruralcolor, 
                 size = 2)
p2R
```


```{r Step3}
p3 <- p2A +annotate("segment", 
                   x = 0.31,
                   xend = 0.31,
                   y = 0.2,
                   size=  1.5,
                   yend = max(twoClass$Income),
                   color = Urbancolor )+
  labs(title = "Third node", 
           subtitle = paste("Rule: ", gsub(",","",toString(myrules[3,2:13])))
       )
p3 

p3A <- p3 + annotate("rect", 
                     xmin= 0.31,
                     xmax = max(twoClass$Education),
                     ymin = 0.2,
                     ymax = max(twoClass$Income),
                     fill = Urbancolor, alpha = 0.2 )
p3A  

# Highlight Misclassified 
misclassU <- twoClass %>%
  filter(Income > 0.2 & Education > 0.13 & Education < .31 
          & classes == "Urban")

misclassR <- twoClass %>%
   filter(Income > 0.2 & Education > 0.13 & Education > .31 
          & classes == "Rural")
    

p3R <- p3A+  geom_point(data = misclassR, 
                 aes(x= Education, y = Income), 
                 colour = Ruralcolor, 
                 size = 2)
p3R
     
p3U <- p3R + geom_point(data = misclassU, 
                 aes(x= Education, y = Income), 
                 colour = Urbancolor, 
                 size = 2)
p3U

```


```{r Step4}
p4 <- p3A +annotate("segment", 
                   x = 0.13,
                   xend = 0.31,
                    y = 0.29,
                   yend = 0.29,
                   size = 1.5,
                   color = Ruralcolor ) +
   labs(title = " Fourth node rule", 
           subtitle = paste("Rule: ", gsub(",","",toString(myrules[2,2:13])))
           )
p4 

p4A <- p4 + annotate("rect", 
                     xmin= 0.13,
                     xmax = 0.31,
                     ymin = 0.29,
                     ymax = max(twoClass$Income),
                     fill = Urbancolor, alpha = 0.2 )
p4A

p4B <- p4A + annotate("rect", 
                        xmin= 0.13,
                        xmax = 0.31,
                        ymin = 0.2,
                        ymax = 0.29,
                        fill = Ruralcolor, alpha = 0.2 )

p4B

# Highlight Misclassified 
misclassU <- twoClass %>%
  filter(Income > 0.2 & Education > 0.13 & Education < .31 
         & Income < 0.29 
         & classes == "Urban")

misclassR <- twoClass %>%
    filter(Income > 0.2 & Education > 0.13 & Education < .31 
         & Income > 0.29 
          & classes == "Rural")
    

p4R <- p4B+  geom_point(data = misclassR, 
                 aes(x= Education, y = Income), 
                 colour = Ruralcolor, 
                 size = 2)
p4R
     
p4U <- p4R + geom_point(data = misclassU, 
                 aes(x= Education, y = Income), 
                 colour = Urbancolor, 
                 size = 2)
p4U


```



# Wrap-up
- Decision tree can be quite easy to interpret








